---
title: "Introduction"
subtitle: "Course overview and expectations"
author: "Trang Tran"
date: "2026-09-22"
format:
  revealjs:
    self-contained: true
    theme: simple
    slide-number: true
    toc: false
    code-line-numbers: false
    #footer: "© Trang Tran: Japan in the World Economy"
    chalkboard: false
    pdf-export: true
    incremental: true
    controls: true
    hash: true
draft: true
output-dir: ../docs/slides
---

## Self-introduction

- Assistant Professor, Department of System Innovation, University of Tokyo
  - International economics
  - Maritime economics

- Why am I here?
  - Buyout system of the Grant-in-Aid for Scientific Research (S)
  - [“Transformation of the Global Value Chain and the
Construction of a New International Economic Order”](
https://www.jsps.go.jp/file/storage/kaken_12_g1990/r_6_en_24h00014.pdf)

---

## Course Content

- What this course covers (trade, ODA, FDI, exchange rates)
- Why Japan? A major economy in the world
- How we’ll learn: lectures, discussion, suggested readings

---

## Expected Outcomes

- Explain Japan’s role in the global economy.

- Assess Japan’s position in global value chains and its competitiveness.

- Formulate evidence-based perspectives on Japan and current global economic issues.

---

## Course Map

- Week 1: Introduction (9/22)
- Weeks 2–6: Global value chain
- **Week 7: Group presentation workshop (11/10)**
- Weeks 8-12: ODA, FDI, Exchange Rates 
- **Week 13: Final exam (12/22)**
- **Make-up exam (1/19)**


---

## International Trade

- Japan's Comparative Advantages

- Measurement of Global Value Chains

- Enablers of Global Value Chains  

- Japan and ASEAN 

- Geopolitics and Climate Change

---

## Selected topics on Japanese economy

- Official Development Assistance

- Cross-border investment

- Balance of Payments

- Exchange Rates  

- Demographic Challenges and Digitization

---


## References

- Takatoshi Ito & Takeo Hoshi, 2020, The Japanese Economy (2nd ed.), The MIT Press 
- Krugman, P. R., Obstfeld, M., & Melitz, M. J. (2023). International economics: Theory and policy (12th ed.). Pearson.
- 石川城太・椋寛・菊地徹『国際経済学をつかむ』第2版有斐閣

---

## How We Work

- Class Structure: 20 min short tests + 50 min lecture  + 5 min break + 30 min group discussions

- Assessment
  - **Participant**: 10%
  - **Short tests:** 20%
  - **Group presentation workshop:** 30%  
  - **Final exam:** 40%

---

- Voluntary and beneficial remark: one or two pts (maximum 10 pts)
- Points deduction: 5 pts (absent), 3 pts (late)
  - Exceptions include illness or delays in public transportation but with proof is needed
  - If you cheat, you will be kicked out

---


## Group discussion (first half of the course)

- 7 groups of 5~6 students: register your members by the end of today
- Choose one research question related to the topic of global value chain in two weeks
- Discuss in class
- Presentation (11/10: 10 minutes ~ 5 slides): title, research question + background, evidence, policy implication, conclusion
- Submit of individual report (11/24: 1 A4 page): similar to presentation + what you learn from other groups

---

### Group discussion (second half of the course)

- Randomly chosen group
- Assigned topics
- Discuss in class

---

## Classroom management:
- “respon” application: https://univ-gakushuin.respon.jp

- Homework: 
  - Prepare for short tests and discussions.

- Office hour: by appointment

---

## Stats of year 2025
- No of registration: 104
- No of final exam takers: 93
- No of credit eaners: 89
  - S: 4
  - A: 30
  - B: 36
  - C: 19
  
---

## Your takeaways

- What interests you in this class ?
- What is your expectation in this class ?

::: notes
Questionnaire 
:::

---

## Japan at a Glance

- High-income
- Economic growth model
- Safe-haven currency
- Aging economy 

::: notes
Use this as a conversation starter: ask for students’ priors on Japan’s role.
:::

---

## High-income

```{r}
#| echo: false
library(data.table)
library(ggpubr)
library(scales)
path = function(x){paste0("../data/01 intro/",x)}
gdppc = fread(path("gdppc.csv"))
pwt = fread(path("growth.csv"))
pwt_trade = fread(path("trade.csv"))
va = fread(path("va.csv"))
unpop = fread(path("pop.csv"))
imfd = fread(path("imf.csv"))

regs = unique(gdppc[variable=="GDPpc",.(V1=median(value)),
                    by=region][order(-V1),region])
regs = c("Japan",regs[regs!="Japan"])
regs_col = c("black",get_palette(palette = "npg",length(regs)-1))
gdppc[,region:=factor(region,levels = regs)]
ggline(gdppc,
       x = "date",y="value",ylab="",xlab="",
       palette = regs_col,
       color = "region",
       title = "GDP per capita (constant 2011 USD)",
       caption = "Source: Madison Project (2023).\n Western Offshoots include Australia, Canada, New Zealand, United States.") +
  guides(color = guide_legend(nrow = 2)) +
  scale_y_continuous(labels = label_number(
    scale = 1/10^3, suffix = "K")) +
  theme_pubclean() 
```
::: notes
- Japan economic growth: rapid 1950s~1960s, stagnant 1970s~1980s, declining 1990s~ 
- Many Asian repeat rapid growth in the 1990s
:::

---

## Economic growth model (input side)

```{r}
ctys = c("JPN","USA","CHN")
ctys_col = c("black",get_palette(palette = "npg",2))
pwt[,ISOcode:=factor(ISOcode,levels = ctys)]

ggarrange(
  ggline(pwt[grepl("service|persons|hours",Variablename2)],
       x="date",y="gyoy",color="ISOcode",palette = ctys_col,
       xlab = "",ylab = "",
       facet.by = "Variablename2",
       numeric.x.axis = T,
       title = "Annual Growth (%)") +
  guides(color = guide_legend(title = "Country")) +
  theme_pubclean(),
  ggline(pwt[grepl("Human|TFP",Variablename2)],
       x="date",y="gyoy",color="ISOcode",palette = ctys_col,
       xlab = "",ylab = "",
       facet.by = "Variablename2",
       numeric.x.axis = T,
       caption = "Source: Penn World Table 10.1") +
  guides(color = guide_legend(title = "Country")) +
  theme_pubclean(),
  common.legend = T,
  nrow = 2
)
```

::: notes
- Growth drivers: capital, total factor productivity, human capital: investment-led
:::

---

## Economic growth model (export)

```{r}
pwt_trade[,countrycode:=factor(countrycode,levels = ctys)]

ggline(pwt_trade[type=="x"],
       x="year",y="value",color="countrycode",palette = ctys_col,
       xlab = "",ylab = "",
       facet.by = "sector",
       numeric.x.axis = T,
       title = "Export Share in GDP at current PPP",
       caption = "Source: PWT 10.01") +
  guides(color = guide_legend(title = "Country")) +
  theme_pubclean()
```

---

## Economic growth (sectoral shift)

```{r}
regs = unique(va[,Region])
regs = c("JPN",regs[regs!="JPN"])
regs_col = c("black",get_palette(palette = "npg",length(regs)-1))
va[,Region:=factor(Region,levels = regs)]

ggline(va[grepl("value added",indicator)&date>=1990],
       x = "date",y="value",ylab="",xlab="",
       title = "Value Added as share of GDP",
       palette = regs_col,add = "none",
       color = "Region",facet.by = "indicator",
       caption = "Source: World Bank") +
  guides(color = guide_legend(nrow = 2)) +
  theme_pubclean() 
```

::: notes
- Shift of labor from agricultural sectors to manufacturing and services
:::
---

## Safe-haven currency

```{r}
curr = unique(imfd[,.(V1=sum(value)),
                   by=Currency][order(-V1),Currency])
curr = c("Japanese yen",curr[curr!="Japanese yen"])
curr_col = c("black",get_palette(palette = "default",length(curr)-1))
imfd[,Currency:=factor(Currency,levels = curr)]

ggbarplot(imfd[!is.na(value)],
          title = "The world total of official foreign exchange reserves by currency",
          x="year",xlab = "",
          y="value",ylab = "",
          fill = "Currency",
          palette = curr_col,
          caption = "IMF COFER") +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  guides(fill = guide_legend(nrow = 3))+
  theme_pubclean()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) 
```


---

## Aging economy (population)

```{r}
unpop[,Region:=factor(Region,levels = regs)]
ggarrange(
  ggline(unpop,
       x = "date",y="MedianAgePop",ylab="",xlab="",
       palette = regs_col,
       title = "Median Age",
       color = "Region") +
  geom_vline(xintercept = 2023) +
  guides(color = guide_legend(nrow = 2)) +
  theme_pubclean(),
  ggline(unpop,
       x = "date",y="NatChangeRT",ylab="",xlab="",
       palette = regs_col,add = "none",
       title = "Rate of Natural Change",
       color = "Region",
       caption = "Source: UN Prospect (2024)") +
  geom_vline(xintercept = 2023) +
  guides(color = guide_legend(nrow = 2)) +
  theme_pubclean(),
  common.legend = T,
  align = "h"
)
```

::: notes
- Rapidly aging, highest median age with very low rate of natural change
:::

---

## Aging economy (immigration)

```{r}
# CNMR  Net Migration Rate (per 1,000 population)
# Net Number of Migrants (thousands)
ggarrange(
  ggline(unpop,
       x = "date",y="CNMR",ylab="per 1000 pop",xlab="",
       palette = regs_col,
       title = "Net migration rate",
       color = "Region") +
  geom_vline(xintercept = 2023) +
  guides(color = guide_legend(nrow = 2)) +
  theme_pubclean(),
  ggline(unpop,
       x = "date",y="NetMigrations",ylab="",xlab="",
       palette = regs_col,add = "none",
       title = "Net migrations",
       color = "Region",
       caption = "Source: UN Prospect (2024)") +
  geom_vline(xintercept = 2023) +
  scale_y_continuous(labels = label_number(
    scale = 1/10^3, suffix = "K")) +
  guides(color = guide_legend(nrow = 2)) +
  theme_pubclean(),
  common.legend = T,
  align = "h"
)
```


---
## Suggested Readings
- Ito and Hoshi (2020), chapter 1-3, 12




 
