---
title: "Comparative Advantage: The case of Japan"
subtitle: ""
author: "Trang Tran"
date: "2026-09-29"
format:
  revealjs:
    self-contained: true
    theme: simple
    slide-number: true
    toc: false
    code-line-numbers: false
    #footer: "© Trang Tran: Japan in the World Economy"
    chalkboard: false
    pdf-export: true
    incremental: true
    controls: true
    hash: true
draft: true
output-dir: ../docs/slides
---

## Introduction

- Japan has capital, labor, and skilled labor 

- In principle, it could grow more of its food

- Instead it imports food and exports machinery

- Why?

## Why countries trade?

- Comparative advantage: ***difference in resources or productivity***

- Economic of scales: similar countries trade if production get cheaper at larger scale

## Absolute vs. Comparative advantage

- Being better at producing everything is absolute advantage: does not determine what gets traded.

- Comparative advantage is about relative opportunity cost: what you give up to produce more unit of something.

- Even a country with no absolute advantage in anything still has a comparative advantage in something.


## Explaining Japan's trade patterns

- Productivity/technology differences

- Short-run immobility

- Capital vs. labor abundance

## Productivity differences

- Core idea: countries export the good where their opportunity cost is lowest.

- Japan's technology edge in precision manufacturing lowers the opportunity cost of producing autos and electronics relative to its trading partners.

- In agriculture, Japan has higher opportunity costs, which driving the import patterns

## Short-run immobility 

- Core idea: some factors (land, sector-specific capital) can't quickly move between industries.

- Japan's farmland and farmers can't retrain into auto manufacturing overnight.

- Despite comparative disadvantage in agriculture, Japan has long-standing  protection policy.

## Factor abundance

- Core idea: countries export goods that intensively use their abundance factor.

- Japan is capital- and technology-abundant relative to land and labor.

- Japan export capital-intensive goods (car, electronics) and imports land/labor-intensive goods (food, raw materials)


## Which lens fits Japan best?

- Exports where relative productivity is highest

- Protection of agriculture despite comparative disadvantage

- Exports capital-intensive goods, imports land/labor-intensive goods

---

## What drives comparative advantage?

::: {.smaller}
Kowalski (2011, OECD): 55 OECD and emerging economies.

:::: {.columns}

::: {.column width="54%"}
**Convergence across countries**

- Physical capital
- Education (tertiary, avg. schooling)
- Primary energy supply
- Financial development (credit)
:::

::: {.column width="44%"}
**Divergence across countries**

- Secondary education
- Regulatory quality, rule of law, corruption
- Import tariffs
:::
::: 
:::

---

## Convergence differs by trade relationship

- North-North (within OECD):	became more homogeneous on most sources

- South-South (within non-OECD): gap remains for capital-to-labor ratio and regulatory indicators but narrowed for education, energy supply, financial development.
 
- North-South (OECD vs. non-OECD): 	Gap widened for physical capital, credit availability, regulatory quality but narrowed for human capital

---

## Revealed comparative advantages

The Balassa index (Balassa, 1965): how much more (or less) a country exports of a good, relative to its share of world trade.

$$
RCA = \frac{\text{country's export share of good X}}
{\text{world's export share of good X}}
$$

- RCA > 1: the country exports proportionally more of that good than the world average

::: notes  
- effective only within same country among products or same product among countries: higher value indicates a stronger export performance, higher degree of export specialization, and a higher degree of revealed comparative advantage
:::  

---

## Questions

1. In what sectors does Japan have high RCA?

2. In what sectors does Japan have low RCA?

## Japan RCA in 2019

```{r}
#| echo: false
library(data.table)
library(ggpubr)
library(ggrepel)

dtcaw_tiva = fread("../data/02 ca/rca_jpn_tiva.csv")

ggscatter(
  dtcaw_tiva,
  x = "export_share",
  y = "traditional",
  size = "export_share",
  color = "traditional",
  xlab = "Japan's Export Share",
  ylab = "RCA",
  title = "Japan: RCA vs. Export Share, 2019",
  subtitle = "Bubble size reflects share of total exports",
  caption = "Source: TIVA from WITS",
  alpha = 0.75
) +
  scale_size_continuous(range = c(2, 12)) +
  scale_color_gradient2(
    midpoint = 1, low = "firebrick", mid = "grey70", high = "steelblue"
  ) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey40", linewidth = 0.4) +
  geom_text_repel(
    aes(label = label_sect),
    size = 3.3,
    fontface = "bold",
    max.overlaps = 20,
    box.padding = 0.5,
    segment.color = "grey60",
    segment.size = 0.3
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "grey30", size = 11),
    plot.caption = element_text(color = "grey50", size = 8),
    panel.grid.major = element_line(color = "grey92", linewidth = 0.3),
    axis.title = element_text(size = 11)
  )

```

## RCA of Japan across period

```{r}
#| echo: false
library(data.table)
library(ggpubr)
library(ggrepel)

dtcaw_jpn = fread("../data/02 ca/rca_jpn.csv")

order_levels = unique(dtcaw_jpn[order(-traditional_med),sect_name])
dtcaw_jpn[,sect_name:=factor(sect_name,levels = order_levels)]

ggboxplot(
  dtcaw_jpn,
  x = "sect_name",
  xlab = "Sector",
  y = "traditional",
  ylab = "RCA",
  rotate = TRUE,
  facet.by = "period",
  nrow = 1,
  title = "RCA by Sector",
  subtitle = "Shift toward capital- and technology-intensive exports",
  caption = "Source: WIOD from WITS",
  fill = "period",
  palette = "Blues",
  outlier.shape = 21,
  outlier.size = 1.5
) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey40", linewidth = 0.4) +
  theme(
    strip.text = element_text(size = 10, face = "bold"),
    strip.background = element_rect(fill = "grey92", color = NA),
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "grey30", size = 11),
    plot.caption = element_text(color = "grey50", size = 8),
    axis.text.y = element_text(size = 9),
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.3),
    panel.spacing = unit(1, "lines")
  )

```

::: notes
- 1965–1974: RCA concentrated in labor-intensive, lower-technology sectors
-- Highest: Textiles, Textile, Leather and Footwear
-- Also elevated: Basic Metals and Fabricated Metal; Machinery, nec.

- 1975–1984: Transition period
-- Textiles RCA begins declining
-- Transport Equipment, Machinery, and Electrical/Optical Equipment start rising

- 1985–2000: RCA concentrated in capital- and technology-intensive sectors
-- Highest: Transport Equipment; Electrical and Optical Equipment; Machinery, nec.
-- Textiles falls to the bottom of the distribution
:::

## RCA: Japan and Others

```{r}
#| echo: false
dtcaw_reg = fread("../data/02 ca/rca_reg.csv")

jpn_vals <- dtcaw_reg[region == "JPN",
                        .(traditional_jpn = traditional), 
                        by = .(sect_name, t)]
usa_vals <- dtcaw_reg[region == "USA",
                        .(traditional_usa = traditional), 
                        by = .(sect_name, t)]
chn_vals <- dtcaw_reg[region == "CHN",
                     .(traditional_chn= traditional), 
                     by = .(sect_name, t)]

ggdensity(
  dtcaw_reg,
  x = "traditional",
  facet.by = c("t", "sect_name"),
  ncol = 3,
  fill = "grey85",
  color = "grey50",
  alpha = 0.6
) +
  geom_vline(
    data = jpn_vals,
    aes(xintercept = traditional_jpn, color = "Japan"),
    linetype = "dashed", linewidth = 0.9
  ) +
  geom_vline(
    data = usa_vals,
    aes(xintercept = traditional_usa, color = "United States"),
    linetype = "dashed", linewidth = 0.9
  ) +
  geom_vline(
    data = chn_vals,
    aes(xintercept = traditional_chn, color = "China"),
    linetype = "dashed", linewidth = 0.9
  ) +
  scale_color_manual(
    name = "Country",
    values = c("Japan" = "#D7263D", "United States" = "#1B6CA8", "China" = "#2E8B57")
  ) +
  guides(color = guide_legend(override.aes = list(linetype = "dashed", linewidth = 1.1))) +
  xlab("RCA") +
  ylab("Density") +
  labs(
    title = "RCA Distribution Across Countries",
    subtitle = "Japan vs. United States vs. China, by sector and period",
    caption = "Source: WIOD from WITS"
  ) +
  theme(
    strip.text = element_text(size = 7, face = "bold"),
    strip.background = element_rect(fill = "grey92", color = NA),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "grey30", size = 11),
    plot.caption = element_text(color = "grey50", size = 8),
    panel.grid.major = element_line(color = "grey93", linewidth = 0.3),
    panel.spacing = unit(0.8, "lines"),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

```
  
---

## Limits of RCA as a measure

- RCA reflects actual trade flows, which are already shaped by tariffs, subsidies, and protection, not a "pure" underlying advantage.

- Global value chains: a "Japanese" car may contain parts made in many countries, blurring what RCA is really measuring.

- Categorization-sensitive: results can shift depending on how narrowly or broadly a "sector" is defined.

---

## Key takeaways

- Japan's trade pattern can be explained by theories on comparative advantages.

- RCA gives us an empirical way to confirm the pattern the theories predict.

- RCA has its limits

---
## Suggested Readings
<div style="font-size:50%">
- Balassa, B. (1965). Trade Liberalisation and “Revealed” Comparative Advantage. The Manchester School, 33(2), 99–123. https://doi.org/10.1111/j.1467-9957.1965.tb00050.x
- OECD (2011), Globalisation, Comparative Advantage and the Changing Dynamics of Trade, OECD Publishing, Paris, https://doi.org/10.1787/9789264113084-en.



</div>
