---
title: "Measurement of Global Value Chains"
subtitle: ""
author: "Trang Tran"
date: "2025-09-29"
format:
  revealjs:
    self-contained: true
    theme: simple
    slide-number: true
    toc: false
    code-line-numbers: false
    #footer: "© Trang Tran: Japan in the World Economy"
    chalkboard: false
    pdf-export: true
    incremental: true
    controls: true
    hash: true
draft: true
output-dir: ../docs/slides
---

## Introduction

- Trade in tasks
- Trade in value added
- Japanese comparative advantage: revisited
- Case study: Motor Vehicles, Electronics, Chemicals

  
## Trade in Tasks

- Traditional trade theory: countries exchange **complete, finished goods**

- Grossman & Rossi-Hansberg (2008): production is unbundled into discrete **tasks** — design, component sourcing, fabrication, assembly, testing, marketing

- Comparative advantage now applies **task-by-task**, not product-by-product

- A task is offshored when it can be performed abroad more cheaply, net of coordination and trade costs

::: notes
- WTO and JETRO-IDE (2011) reframe East Asian trade as evolving "from trade in goods to trade in tasks"
- Explains the rise of intra-industry, intra-firm trade in intermediate goods
- A country's export mix reflects which tasks it specializes in, not which whole product it exports
:::

## An example

```{r}
library(data.table)
library(ggpubr)
library(ggrepel)
iphone = fread("../data/03 gvc/gvc_iphone.csv")
iphone[,rid:=1:.N]
iphone[,label:=ifelse(rid<9|rid>15,Manufaturer_Component,"")]
ggpie(iphone,
      x = "Cost USD",                 
      label = "label", 
      fill = "Country",
      lab.adjust = 0.3,
      label_percent = TRUE,
      title = "Manufacturing Cost of iphone 3G",
      caption = "Source: Xing and Detert (2010) ") +
  theme(legend.position = "right")
```


## The Smile Curve

![Global Value Chain  
_Source: Baldwin and Ito (2021)_](image/gvc_smilecurve.jpg)

::: notes
- Introduced by Acer founder and CEO Stan Shih in the early 1990s: - the early and late stages of the supply chains generate higher value added
- The share of value added in manufactured products is shifting from the fabrication stages to pre- and post-fabrication services. 
  - reclassification: from in-house to outsourced -> factory-less manufacturer
  - task-composition shifts: more logistics, transportation, communication to move intermediates or more software/design etc final goods
  - task–relative price shifts: off-shoring to reduce the relative price of the off-shored tasks.

::: 


## Trade in Value Added

- Gross trade statistics **double-count** intermediate inputs that cross borders multiple times before reaching the final consumer

- The Trade in Value Added (TiVA) approach reallocates exports to the country that actually created the value:
  - **Domestic Value Added (DVA)**: value added by the exporting country itself
  - **Foreign Value Added (FVA)**: value added by other countries, embodied in imported inputs

::: notes
- Gross exports overstate bilateral imbalances (e.g., the US–China trade deficit narrows substantially once measured in value-added terms)
- Value-added trade better reflects which country actually benefits from a given trade flow
:::


## Japan comparative advantage: revisited

```{r}
dtcaw_jpn = fread("../data/02 ca/rca_jpn.csv")
dtcaw_jpn_s = dtcaw_jpn[t%in%c(1965,2000)]
ggscatter(dtcaw_jpn_s,
          title = "Revealed Comparative Advantage for Japan",
          caption = "Data Source: WITS, WIOD, TIVA",
              color = "sect_group",
          legend="none",
          x="traditional",y="gvc",
          facet.by = "t",
          xlab="Gross Trade",
          ylab="Trade in value added") +
  geom_abline(intercept = 0, slope = 1, color = "red",
              linetype = "dotdash") +
  geom_hline(yintercept = 1, linetype = "dotdash") +
  geom_vline(xintercept = 1, linetype = "dotdash") +
    geom_text_repel(aes(x = traditional, y = gvc, label = sect_name, 
                      color = sect_group),
                    data = dtcaw_jpn_s, size = 2,
                   show.legend = FALSE)

```
  
::: notes  
- RCA_Ai = share of export of industry i in country A export/ share of global export of industry in global export (Balassa, 1965)
- effective only within same country among products or same product among countries: higher value indicates a stronger export performance, higher degree of export specialization, and a higher degree of revealed comparative advantage
- CA emphasize relative differences in productivity between countries as the reason for international trade
:::  

---
```{r}
#| fig-height: 9
#| fig-width: 12
dtcaw_reg = fread("../data/02 ca/rca_reg.csv")
regss = unique(dtcaw_reg[,region])
dtcaw_reg[,region:=factor(region,levels = regss)]
dtcaw_reg_s = dtcaw_reg[grepl("Electrical",sect_name)&t%in%c(1970,2000)]
ggscatter(dtcaw_reg_s,
    title = "Revealed Comparative Advantage for Electrical Equipment",
          caption = "Data Source: WITS, TIVA",
          x="traditional",y="gvc",
          #color = "sect_group",
    xlab="Gross Trade",ylab =" Trade in value added",
          legend="none",
facet.by = "t",nrow=1) +
  geom_abline(intercept = 0, slope = 1, 
              linetype = "dotdash") +
  geom_hline(yintercept = 1, linetype = "dotdash") +
  geom_vline(xintercept = 1, linetype = "dotdash") +
    geom_text_repel(aes(x = traditional, y = gvc, label = region, 
                      #color = sect_group
                      ),
                    data = dtcaw_reg_s, size = 3,max.overlaps = 20,
                   show.legend = FALSE)
```

::: notes  
- the highest degree of RCA variation: primary and unskilled labour-intensive product
- less variation in technology and human capital-intensive product categories
:::  

## Key Features of Global Value Chains 

- Fragmentation of Production: across multiple countries for cost structures, labor markets, and expertise

- Specialization: specific segments of the chain (R&D, design, manufacturing, logistics )
   
- Interconnection: trade, investment, and technology

- Value creation: value added chain from materials to final goods

- Global distribution: multiple markets

## Measuring Global Value Chains (1)

- GVC participation decomposes a country's **gross exports** by the source and destination of value added.

**Backward linkage (backward participation)**

$$ \text{Backward} = \frac{FVA}{\text{Gross Exports}} \times 100 $$

- $FVA$ = foreign value added embodied in a country's own exports
- Captures how much a country **relies on imported inputs** to produce what it exports

## Measuring Global Value Chains (2)

**Forward linkage (forward participation)**

$$ \text{Forward} = \frac{DVX}{\text{Gross Exports}} \times 100 $$

- $DVX$ = domestic value added exported *indirectly*, i.e. used as an input in a third country's exports
- Captures how much of a country's own value added is **used by others** further downstream

- Total GVC participation = Backward + Forward

::: notes
- Based on the gross export decomposition in Koopman, Wang & Wei (2014), built from international input–output tables
- Small, open economies (e.g., ASEAN members) tend to show higher backward participation — they import more of what they need to export
- Countries positioned at the start of chains (raw materials, core components, R&D) tend to show higher forward participation
- Rule of thumb: Backward ↑ when a country imports a lot of intermediates; Forward ↑ when a country's intermediates are re-exported by others
:::

## Finer measurement

- GVC-related trade: cross border at least twice
  - **Pure backward participation**: the use of foreign inputs in exports to the final destination (chain end)
  - **Pure forward participation**: the use of domestically produced intermediates in third-country exports (chain start)
  - **Mixed participation**: the use of foreign intermediates in third-country exports

::: notes
- Upstream (high value): design, R&D, core components, precise machinery
- Midstream: capital goods exports, management of production networks
- Downstream: limited (assembly shifted to China/ASEAN), but could be different for chemicals (pharmaceutical synthesis)
::: 


## Why GVCs are Important

```{r}
dtw = fread("../data/03 gvc/gvc_share.csv")
ggarrange(
  ggpie(dtw[t==2020],
        x="share",
        color="variable",
        fill="variable",
        legend="none",
        label="variable",
        title = "2020") +
    theme(plot.margin = unit(c(1,1,1,1), "cm")),
  ggpie(dtw[t==1970],
        x="share",
        color="variable",
        fill="variable",
        legend="none",
        label="variable",
        title = "1970",
        caption = "Source: WITS, WIOD, TIVA")+
    theme(plot.margin = unit(c(1,2,1,1), "cm")),
  align = "h"
) %>% 
  annotate_figure(
    top  = text_grob("World Trade", face = "bold", size = 14))
```

::: notes
- Only one fourth trade are GVC-related in 1970
- Over 50% of world trade occurs within GVCs in 2020 
- Enables technology transfer and knowledge spillovers.  
- Entry point for developing countries into global trade.  
- Also increases vulnerability to shocks (pandemic, natural disasters).  
:::


## Why GVCs Matter for Japan

```{r}
dt_gvc = fread("../data/03 gvc/gvc_JPN.csv")
ggbarplot(dt_gvc,
          x="t",xlab = "",y="share",
          fill="variable",
          color = "variable",
          facet.by = "type",
          caption = "Data Source: WITS, WIOD, TIVA",
          title="Japan trade") +
  #theme(plot.margin = unit(c(1,1,1,1), "cm")) +
  guides(fill=guide_legend(nrow=2),color=guide_legend(nrow=2))
```
::: notes
- GVC-related trade increase from about 20% in 1970 to 40% in 2020
- Further separation of production chain with assembly move to China in 1990s, and then new ASEAN3
:::

## Japan's in GVC Participation

```{r}
dt_index = fread("../data/03 gvc/gvc_index.csv")
ggarrange(
  ggbarplot(dt_index[t==1970],
            x="RegionRank",y="share",
            facet.by = "t",
            color = "variable",
            fill = "variable") +
    theme(axis.text.x = element_text(angle = 90)),
  ggbarplot(dt_index[t==2020],
            x="RegionRank",y="share",
            facet.by = "t",
            color = "variable",
            fill = "variable") +
    theme(axis.text.x = element_text(angle = 90)),
  common.legend = T
) %>% annotate_figure(
  bottom = "Data Source: WITS"
)
```

::: notes
- Small open economies source more inputs from abroad in GVCs than large countries: more backward participation
:::

## Japanese import and export structure

```{r}
dt_gvc_reg = fread("../data/03 gvc/gvc_reg.csv")
ggbarplot(dt_gvc_reg,
          caption = "Data Source: WITS, TIVA",
          x="region",
          xlab="",
          ylab = "billion USD",
          y="value2",
          fill="variable",
          color = "variable",
          facet.by = c("sect_group"),
          #title="Japan trade",
          rotate =T) +
  #theme(plot.margin = unit(c(1,1,1,1), "cm")) +
  geom_hline(yintercept = 0,color = "black", linetype = "dotdash")
```

::: notes
- Import energy, food, raw materials mostly traditional trade
- Japan trade with the N.America/EU
  - import services
  - export manufacturing goods, surplus with US
- Japan trade with China/ASEAN
  - Imports consumer electronics, textiles
  - Exports high-tech machinery, components
  - "Task sharing" production networks
  
- services (WTO GATS): production happens with consumption
  - cross-border supply 
  - consumption abroad (tourism, education)
  - commercial presence (branch of bank)
  - movement of natural persons (Filipino nurse contract in Japan: not migration)

:::


## Regional Hubs

![Global value chain and regional hubs
_Source: Suvannaphakdy and Pham (2023) based on Abiad et al. (2018)_](image/gvc_network.png)

## Case: Vehicles

```{r}
#| fig-height: 6
#| fig-width: 12
dtback = fread("../data/03 gvc/gvc_backward.csv")
cregss = c("JPN","CHN","KOR","ASEAN","Europe","N.America","ROW")
dtback[,exporter:=factor(exporter,levels = cregss)]
cregss2 = c("JCK+ASEAN","Europe","N.America","ROW")
dtback[,importer:=factor(importer,levels = cregss2)]

ggbarplot(dtback[grepl("Motor vehicles",sect_name)],
          x="exporter",y="value",
          ylab = "GVC Trade (mil USD)",
          facet.by = c("importer","t"),
          color="variable",
          fill = "variable") +
  theme(axis.text.x = element_text(angle = 90)) 
```

::: notes
- regional factory: N.America, Europe, Asia
- transportation cost:  costly to move cars
- regional specialization
::: 

---

![Japanese Autos by 1990s 
_Source: Kinoshita (2004) based on Yamaguchi (2003),  Honda Motor Co., Ltd (2003)_ ](image/gvc_veh.png)

::: notes
- Japan adopt a strategy of concentrating production of similar products in a single country to attain economies of scale
- Shift of assembly type operations from ASEAN to China to utilize low-cost labor 
- Shift from China to CLMV for labor-intensive production
- ASEAN assembles vehicles
- "Factory Asia"
::: 


## Case: Electronics

```{r}
#| fig-height: 6
#| fig-width: 12
ggbarplot(dtback[grepl("electronic",sect_name)],
          x="exporter",y="value",
          ylab = "GVC related trade (mil USD)",
          facet.by = c("importer","t"),
          color="variable",
          fill = "variable") +
  theme(axis.text.x = element_text(angle = 90))
```

::: notes
- more global value chain
  - Chips & components designed in Japan
  - Manufactured/assembled in ASEAN or China
  - Finished products shipped worldwide
  
:::

## Case: Chemicals

```{r}
#| fig-height: 6
#| fig-width: 12
ggbarplot(dtback[grepl("Chemical",sect_name)],
          x="exporter",y="share",
          ylab = "GVC related trade (mil USD)",
          facet.by = c("importer","t"),
          color="variable",
          fill = "variable") +
  theme(axis.text.x = element_text(angle = 90))
```

::: notes
- less outsources than electronics or vehicles
- regional clusters are larger
- some degree of gvc participation with some high import content
:::

## Case: Agricultural products

```{r}
#| fig-height: 6
#| fig-width: 12
ggbarplot(dtback[grepl("Agriculture",sect_name)],
          x="exporter",y="share",
          ylab = "GVC related trade (mil USD)",
          facet.by = c("importer","t"),
          color="variable",
          fill = "variable") +
  theme(axis.text.x = element_text(angle = 90))
```

::: notes
- high internally input
- sourced mostly domestically, but some increasing trend in import content as well
:::

## Challenges

- Demographics  
- Geopolitics  
- Sustainability  

## Conclusion

- Japan remains a **vital player and architect** of GVCs.  
- Future depends on **innovation, resilience, and green supply chains**.  

---
## Suggested Readings
<div style="font-size:50%">
- Baldwin, R., & Ito, T. (2021). The smile curve: Evolving sources of value added in manufacturing. Canadian Journal of Economics/Revue Canadienne d’économique, 54(4), 1842–1880. 
- Elms, D. K., & Low, P. (2013). Global value chains in a changing world. World trade organization.
- Kinoshita, T., Kishida, H., & Amemiya, A. (2004). Strategy of Japanese manufacturers in ASEAN and China in light of emerging FTA initiatives.
- OECD. (2011). Globalisation, comparative advantage and the changing dynamics of trade (Online-Ausg). OECD.
- Suvannaphakdy, S., & Pham, T. P. T. (2023). GVC Reconfiguration. ISEAS–Yusof Ishak Institute Singapore.
- Todo, Y., & Inoue, H. (2021). Geographic Diversification of the Supply Chains of Japanese Firms. Asian Economic Policy Review, 16(2), 304–322
- WTO and JETRO-IDE (2011), Trade patterns and global value chains in East Asia: from trade in goods to trade in tasks. 
- Xing, Y., and N. Detert (2010). How the iPhone Widens the United States Trade Deficit with the People’s Republic of China. ADBI Working Paper 257. Tokyo: Asian Development Bank.

## Data
Mancini, M., & Taglioni, D. (2025). Economic consequences of trade and global value chain integration: A measurement perspective. The World Bank Economic Review, https://doi.org/10.1093/wber/lhaf017.

</div>







 
