---
title: "Measurement of Global Value Chains"
subtitle: ""
author: "Trang Tran"
date: "2025-10-06"
format:
  revealjs:
    self-contained: true
    theme: simple
    slide-number: true
    toc: false
    code-line-numbers: false
    #footer: "© Trang Tran: Japan in the World Economy"
    chalkboard: false
    pdf-export: true
    incremental: true
    controls: true
    hash: true
draft: true
output-dir: ../docs/slides
---

## Introduction

- Trade in tasks
- Trade in value added
- Japanese comparative advantage: revisited
- Case study: Motor Vehicles, Electronics, Chemicals

  
## Trade in Tasks

- Traditional trade theory: countries exchange **complete, finished goods**

- Grossman & Rossi-Hansberg (2008): production is unbundled into discrete **tasks** — design, component sourcing, fabrication, assembly, testing, marketing

- Comparative advantage now applies **task-by-task**, not product-by-product

- A task is offshored when it can be performed abroad more cheaply, net of coordination and trade costs

::: notes
- WTO and JETRO-IDE (2011) reframe East Asian trade as evolving "from trade in goods to trade in tasks"
- Explains the rise of intra-industry, intra-firm trade in intermediate goods
- A country's export mix reflects which tasks it specializes in, not which whole product it exports
:::

## An example

```{r}
#| echo: false
#| warning: false
library(data.table)
library(ggpubr)
library(ggrepel)
library(patchwork)
iphone = fread("../data/03 gvc/gvc_iphone.csv")
iphone[,rid:=1:.N]
iphone[,label:=ifelse(rid<9|rid>15,Manufaturer_Component,"")]
ggpie(iphone,
      x = "Cost USD",                 
      label = "label", 
      fill = "Country",
      lab.adjust = 0.3,
      label_percent = TRUE,
      title = "Manufacturing Cost of iphone 3G",
      caption = "Source: Xing and Detert (2010) ") +
  theme(legend.position = "right")
```


## The Smile Curve

```{r}
#| echo: false
#| message: false
#| warning: false

# ============================================================
# 1. Function to create a smooth curve from control points
# ============================================================

make_curve <- function(x, y, n = 200) {
  
  s <- spline(x, y, n = n)
  
  data.frame(
    x = s$x,
    y = s$y
  )
}


# ============================================================
# 2. Function to draw one smile-curve panel
# ============================================================

smile_plot <- function(curve_new, curve_old,
                       points_new, points_old,
                       x_labels,
                       title,
                       y_title,
                       x_title,
                       new_label_pos,
                       old_label_pos) {
  
  ggplot() +
    
    # Axes
    geom_segment(
      aes(x = 0, y = 0, xend = 1.02, yend = 0),
      arrow = arrow(length = unit(0.18, "cm")),
      linewidth = 0.8
    ) +
    
    geom_segment(
      aes(x = 0, y = 0, xend = 0, yend = 0.82),
      arrow = arrow(length = unit(0.18, "cm")),
      linewidth = 0.8
    ) +
    
    # 21st-century curve
    geom_line(
      data = curve_new,
      aes(x, y),
      linewidth = 0.9
    ) +
    
    # 1970s curve
    geom_line(
      data = curve_old,
      aes(x, y),
      linewidth = 0.8,
      linetype = "dashed"
    ) +
    
    # Points
    geom_point(
      data = points_new,
      aes(x, y),
      shape = 21,
      size = 9,
      stroke = 0.8,
      fill = "white"
    ) +
    
    geom_point(
      data = points_old,
      aes(x, y),
      shape = 21,
      size = 9,
      stroke = 0.8,
      fill = "white"
    ) +
    
    # X-axis category labels
    geom_text(
      data = x_labels,
      aes(x, y, label = label),
      fontface = "italic",
      size = 4.5,
      vjust = 1
    ) +
    
    # Curve labels
    annotate(
      "text",
      x = new_label_pos[1],
      y = new_label_pos[2],
      label = "21st century\nvalue chain",
      hjust = 0,
      size = 4.5
    ) +
    
    annotate(
      "text",
      x = old_label_pos[1],
      y = old_label_pos[2],
      label = "1970s\nvalue chain",
      hjust = 0,
      size = 4.5
    ) +
    
    # Axis titles
    annotate(
      "text",
      x = 0,
      y = 0.87,
      label = y_title,
      size = 4.5
    ) +
    
    annotate(
      "text",
      x = 1.04,
      y = 0,
      label = x_title,
      hjust = 0,
      size = 4.5
    ) +
    
    # Panel title
    annotate(
      "label",
      x = 0.52,
      y = 0.97,
      label = title,
      size = 5.5,
      fill = "white"
    ) +
    
    coord_cartesian(
      xlim = c(-0.08, 1.20),
      ylim = c(-0.15, 1.03),
      clip = "off"
    ) +
    
    theme_void() +
    
    theme(
      plot.margin = margin(10, 25, 30, 25)
    )
}


# ============================================================
# 3. FIRM-LEVEL SMILE CURVE
# ============================================================

# Control points
firm_new_pts <- data.frame(
  x = c(.17, .50, .83),
  y = c(.64, .22, .64)
)

firm_old_pts <- data.frame(
  x = c(.17, .50, .83),
  y = c(.48, .38, .48)
)

# Smooth curves
firm_new <- make_curve(
  firm_new_pts$x,
  firm_new_pts$y
)

firm_old <- make_curve(
  firm_old_pts$x,
  firm_old_pts$y
)

# Stage labels
firm_labels <- data.frame(
  x = c(.17, .50, .83),
  y = c(-.035, -.065, -.035),
  label = c(
    "Pre-fab\nservices",
    "Fabrication",
    "Post-fab\nservices"
  )
)

# Plot
p1 <- smile_plot(
  curve_new = firm_new,
  curve_old = firm_old,
  points_new = firm_new_pts,
  points_old = firm_old_pts,
  x_labels = firm_labels,
  
  title = "Firm-level smile curve",
  y_title = "Stage's share of\nproduct value added",
  x_title = "Stage",
  
  new_label_pos = c(.55, .72),
  old_label_pos = c(.82, .42)
)


# ============================================================
# 4. ECONOMY-WIDE SMILE CURVE
# ============================================================

# Control points
economy_old_pts <- data.frame(
  x = c(.15, .50, .88),
  y = c(.18, .73, .20)
)

economy_new_pts <- data.frame(
  x = c(.15, .50, .88),
  y = c(.20, .50, .42)
)

# Smooth curves
economy_old <- make_curve(
  economy_old_pts$x,
  economy_old_pts$y
)

economy_new <- make_curve(
  economy_new_pts$x,
  economy_new_pts$y
)

# Sector labels
economy_labels <- data.frame(
  x = c(.15, .50, .88),
  y = -.06,
  label = c(
    "Primary",
    "Manufacturing",
    "Services"
  )
)

# Plot
p2 <- smile_plot(
  curve_new = economy_new,
  curve_old = economy_old,
  points_new = economy_new_pts,
  points_old = economy_old_pts,
  x_labels = economy_labels,
  
  title = "Economy-wide smile curve",
  y_title = "Input-sector's share of\nexport value added",
  x_title = "Origin\nsector",
  
  new_label_pos = c(.90, .38),
  old_label_pos = c(.67, .69)
)


# ============================================================
# 5. Combine
# ============================================================

(p1 + p2) +
  plot_annotation(
    caption = "Source: Adapted from Baldwin and Ito (2021)."
  ) &
  theme(
    plot.caption = element_text(
      size = 10,
      hjust = 0
    )
  )

```

::: notes
- Introduced by Acer founder and CEO Stan Shih in the early 1990s: - the early and late stages of the supply chains generate higher value added
- The share of value added in manufactured products is shifting from the fabrication stages to pre- and post-fabrication services. 
  - reclassification: from in-house to outsourced -> factory-less manufacturer
  - task-composition shifts: more logistics, transportation, communication to move intermediates or more software/design etc final goods
  - task–relative price shifts: off-shoring to reduce the relative price of the off-shored tasks.

::: 


## Trade in Value Added

- Gross trade statistics **double-count** intermediate inputs that cross borders multiple times before reaching the final consumer

- The Trade in Value Added (TiVA) approach reallocates exports to the country that actually created the value:
  - **Domestic Value Added (DVA)**: value added by the exporting country itself
  - **Foreign Value Added (FVA)**: value added by other countries, embodied in imported inputs

::: notes
- Gross exports overstate bilateral imbalances (e.g., the US–China trade deficit narrows substantially once measured in value-added terms)
- Value-added trade better reflects which country actually benefits from a given trade flow
:::

## Simple example

```{mermaid}
flowchart LR
    A["Country A<br/>Component<br/><b>$20 VA</b>"] -->
    B["Country B<br/>Adds $30 VA<br/><b>Export = $50</b>"] -->
    C["Country C<br/>Adds $50 VA<br/><b>Final product = $100</b>"] -->
    D["Final consumer"]
```

Gross exports recorded


$$
\text{Gross exports} = 20 + 50 + 100 = \$170
$$

Total value added is

$$
20 + 30 + 50 = \$100
$$


But how do we know that all $20 imported by B is actually used to produce B's exports?

## Input–Output Tables 

$$
\boxed{\text{Who supplies inputs to whom?}}
$$


```{mermaid}
flowchart LR
    A["Domestic<br/>Industry A"] <--> M["Domestic<br/>Manufacturing"]

    I["Imported<br/>inputs"] --> A
    I --> M

    A --> E["Exports"]
    M --> E

    A --> D["Domestic<br/>final demand"]
    M --> D
```

## MRIO Data Used in WITS GVC Indicators

**WITS GVC** combines GVC indicators calculated from several
global input–output databases:

| Database | Countries | Sector | Period
|---|---|---|---|
| Eora26 | 189 | 26| 1990-2015
| WIOD 2016 | 43 + RoW | 56 | 2000-2014
| Long-run WIOD | 25 | 23 | 1965-2000
| OECD TiVA/ICIO |76 | 45 | 1995-2020
| ADB MRIO | 62 | 35 | 2007-2022


## Global Value Chains

GVCs refer to the fragmentation of production across countries, where different stages of producing a good or service are carried out in different countries.

::: columns

::: {.column width="50%"}

**Snake: Sequential production**

```{mermaid}
flowchart LR
    A["Raw<br/>materials"] -->
    B["Parts"] -->
    C["Processing"] -->
    D["Assembly"] -->
    E["Final<br/>product"]
```

:::

::: {.column width="50%"}

**Spider: Assembly of components**

```{mermaid}
flowchart TB
    A["Part A"] --> E["Assembly"]
    B["Part B"] --> E
    C["Part C"] --> E
    D["Part D"] --> E
    E --> F["Final<br/>product"]
```

:::

:::

::: {.footer}
Source: Adapted from Baldwin and Venables (2013)
:::

## Key Features of Global Value Chains 

- Fragmentation of Production: across multiple countries for cost structures, labor markets, and expertise

- Specialization: specific segments of the chain (R&D, design, manufacturing, logistics )
   
- Interconnection: trade, investment, and technology

- Value creation: value added chain from materials to final goods

- Global distribution: multiple markets

## Backward and Forward GVC Linkages

::: columns

::: {.column width="50%"}

### Backward linkage

```{mermaid}
flowchart LR
    F["Foreign<br/>value added"] --> H["FOCAL<br/>COUNTRY"] --> D["Exports"]
```

The country **uses foreign inputs** in its exports.

:::

::: {.column width="50%"}

### Forward linkage

```{mermaid}
flowchart LR
    H["FOCAL<br/>COUNTRY"] --> F["Foreign<br/>production"] --> D["Exports"]
```

The country's value added is **used in other countries' exports**.

:::

:::

## Measuring GVC Participation

**Backward linkage (backward participation)**

$$ \text{Backward} = \frac{FVA}{\text{Gross Exports}} \times 100 $$
**Forward linkage (forward participation)**

$$ \text{Forward} = \frac{DVX}{\text{Gross Exports}} \times 100 $$
**Total GVC participation = Backward + Forward**

::: notes
- Based on the gross export decomposition in Koopman, Wang & Wei (2014), built from international input–output tables
- Small, open economies (e.g., ASEAN members) tend to show higher backward participation. They import more of what they need to export
- Countries positioned at the start of chains (raw materials, core components, R&D) tend to show higher forward participation
- Rule of thumb: Backward ↑ when a country imports a lot of intermediates; Forward ↑ when a country's intermediates are re-exported by others
:::

## But, Where Does This Country Stand?

```{mermaid}
flowchart LR
    A["Country A"] --> B["COUNTRY B"] --> C["Country C"] --> D["Final market"]
```

Country B:

- imports intermediate inputs from Country A
- adds domestic value
- exports intermediates to Country C
- Country C uses them to produce exports

**Question: Is Country B participating through a backward or a forward linkage?**

## Pure and Mixed GVC Participation

Pure backward: Imports FVA → produces exports

```{mermaid}
flowchart LR
    A["Foreign inputs"] --> B["FOCAL COUNTRY"] --> C["Final exports"]
```

Pure forward: Exports DVA → used in foreign exports

```{mermaid}
flowchart LR
    A["FOCAL COUNTRY"] --> B["Foreign production"] --> C["Final exports"]
```

Mixed: Imports FVA and contributes value added to downstream foreign exports

```{mermaid}
flowchart LR
    A["Foreign inputs"] --> B["FOCAL COUNTRY"] --> C["Foreign production"] --> D["Final exports"]
```

## Decompose Gross Trade

- Traditional trade: cross border once
- GVC-related trade: cross border at least twice
  - **Pure backward participation**: the use of foreign inputs in exports to the final destination (chain end)
  - **Pure forward participation**: the use of domestically produced intermediates in third-country exports (chain start)
  - **Mixed participation**: the use of foreign intermediates in third-country exports


## GVC intensity and GVC position

- How strongly is a country or industry involved in GVC production?

$$
\text{GVC Participation}
=
\text{Pure Backward}
+
\text{Pure Forward}
+
\text{Mixed}
$$

- Where does the country or industry tend to operate in the value chain?

$$
Forwardness = \frac{\text{Pure Forward} - \text{Pure Backward}}{\text{Total GVC Participation}}
$$

::: notes
Alternative measures of GVC position include **upstreamness** 
(distance of output from final demand) and **downstreamness** 
(distance from primary inputs/value added). These measures are conceptually 
related to, but distinct from, WITS forwardness.

- Upstream (high value): design, R&D, core components, precise machinery
- Midstream: capital goods exports, management of production networks
- Downstream: limited (assembly shifted to China/ASEAN), but could be different for chemicals (pharmaceutical synthesis)
:::

## Global GVC-related trade

```{r}
#| echo: false
dtw = fread("../data/03 gvc/gvc_share.csv")
ggarrange(
  ggpie(dtw[t==2020],
        x="share",
        color="variable",
        fill="variable",
        legend="none",
        label="variable",
        title = "2020") +
    theme(plot.margin = unit(c(1,1,1,1), "cm")),
  ggpie(dtw[t==1970],
        x="share",
        color="variable",
        fill="variable",
        legend="none",
        label="variable",
        title = "1970",
        caption = "Source: WITS, WIOD, TIVA")+
    theme(plot.margin = unit(c(1,2,1,1), "cm")),
  align = "h"
) %>% 
  annotate_figure(
    top  = text_grob("World Trade", face = "bold", size = 14))
```

::: notes
- Only one fourth trade are GVC-related in 1970
- Over 50% of world trade occurs within GVCs in 2020 
- Enables technology transfer and knowledge spillovers.  
- Entry point for developing countries into global trade.  
- Also increases vulnerability to shocks (pandemic, natural disasters).  
:::

## Regions with high GVC participation

```{r}
#| echo: false
dt_index = fread("../data/03 gvc/gvc_region.csv")
cols = get_palette("default",5)[1:3]
ggarrange(
  ggbarplot(dt_index[t==1970],
            x="RegionRank",y="share",
            facet.by = "t",
            palette = cols,
            color = "type",
            fill = "type") +
    theme(axis.text.x = element_text(angle = 90)),
  ggbarplot(dt_index[t==2020],
            x="RegionRank",y="share",
            facet.by = "t",
            palette = cols,
            color = "type",
            fill = "type") +
    theme(axis.text.x = element_text(angle = 90)),
  common.legend = T
) %>% annotate_figure(
  bottom = "Data Source: WITS"
)
```

::: notes
- Small open economies source more inputs from abroad in GVCs than large countries: more backward participation
:::

## Forwardness Distribution

```{r}
#| echo: false
dt_index_w = fread("../data/03 gvc/gvc_region_forward.csv")

jpn_vals <- dt_index_w[region == "JPN",.(t,forwardness)]
usa_vals <- dt_index_w[region == "USA",.(t,forwardness)]
chn_vals <- dt_index_w[region == "CHN",.(t,forwardness)]

ggdensity(dt_index_w,
          x="forwardness",
          facet.by="t") +
  geom_vline(
    data = jpn_vals,
    aes(xintercept = forwardness, color = "Japan"),
    linetype = "dashed", linewidth = 0.9
  ) +
  geom_vline(
    data = usa_vals,
    aes(xintercept = forwardness, color = "United States"),
    linetype = "dashed", linewidth = 0.9
  ) +
  geom_vline(
    data = chn_vals,
    aes(xintercept = forwardness, color = "China"),
    linetype = "dashed", linewidth = 0.9
  ) +
  scale_color_manual(
    name = "Country",
    values = c("Japan" = "#D7263D", "United States" = "#1B6CA8", "China" = "#2E8B57")
  ) +
  guides(color = guide_legend(override.aes = list(linetype = "dashed", linewidth = 1.1))) +
  xlab("Forwardness") +
  ylab("Density") 
```


## Japan's GVC-related trade share

```{r}
#| echo: false
dt_gvc = fread("../data/03 gvc/gvc_JPN.csv")
ggbarplot(dt_gvc,
          x="t",xlab = "",y="share",
          fill="type",
          color = "type",
          facet.by = "flow",
          caption = "Data Source: WITS, WIOD, TIVA") +  guides(fill=guide_legend(nrow=1),color=guide_legend(nrow=1))
```
::: notes
- GVC-related trade increase from about 20% in 1970 to 40% in 2020
- Further separation of production chain with assembly move to China in 1990s, and then new ASEAN3
:::

## Japan's GVC-related trade share by sector

```{r}
#| echo: false
gvc_jpn_sect = fread("../data/03 gvc/gvc_jpn_sect.csv")
gvc_jpn_sect_s = gvc_jpn_sect[grepl("gvc",type)&!is.na(gvc1970)]
gvc_jpn_sect_s[,label:=fifelse(gvc2000>0.2,sect_name,"")]

ggscatter(gvc_jpn_sect_s,
          caption = "Data Source: WITS, WIOD",
          x="gvc1970",y="gvc2000",
    xlab="GVC share in 1970",
    ylab ="GVC share in 2000",
          legend="none",
facet.by = "type",nrow=1) +
  geom_abline(intercept = 0, slope = 1, 
              linetype = "dotdash") +
  geom_hline(yintercept = 0.6, linetype = "dotdash") +
  geom_vline(xintercept = 0.6, linetype = "dotdash") +
  geom_text_repel(aes(x = gvc1970, 
                      y = gvc2000, 
                      label = label, 
                      ),
                    data = gvc_jpn_sect_s, size = 2,max.overlaps = 20,
                   show.legend = FALSE)
```

## Japan's sector position

```{r}
#| echo: false
gvc_jpn_sect_ww = fread("../data/03 gvc/gvc_jpn_sect_forward.csv")

ggscatter(gvc_jpn_sect_ww,
          caption = "Data Source: WITS, WIOD",
          x="forward1970",y="forward2000",
          xlab="Forwardness in 1970",
          ylab ="Forwardness in 2000",
          legend="none") +
  geom_abline(intercept = 0, slope = 1, 
              linetype = "dotdash") +
  geom_hline(yintercept = 0, linetype = "dotdash") +
  geom_vline(xintercept = 0, linetype = "dotdash") +
  geom_text_repel(aes(x = forward1970, 
                      y = forward2000, 
                      label = sect_name, 
  ),
  data = gvc_jpn_sect_ww, size = 2,max.overlaps = 20,
  show.legend = FALSE)
```


## Four sectors globally

```{r}
#| echo: false
#| fig-height: 6
#| fig-width: 12

dt_sec = fread("../data/03 gvc/gvc_case_index.csv")

p_scatter = ggplot(
  dt_sec[t%in%c(2020)],
  aes(
    x = forwardness,
    y = intensity,
    color = sect_name
  )
) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.6, size = 2) +
  facet_wrap(~t) +
  scale_x_continuous(
    limits = c(-1, 1),
    breaks = seq(-1, 1, .5)
  ) +
  labs(
    x = "GVC Position (Forwardness)",
    y = "GVC Intensity",
    color = "Sector"
  ) +
  stat_density2d(
    aes(x = forwardness, y = intensity, color =  sect_name),
    contour = TRUE,
    linewidth = 0.3
  ) +
  theme_minimal()

p_x <- ggplot(dt_sec[t %in% c(2020)],
              aes(forwardness, color = sect_name, fill = sect_name)) +
  geom_density(alpha = 0.4) +
  theme_void() +
  theme(legend.position = "none")

p_y <- ggplot(dt_sec[t %in% c(2020)],
              aes(intensity, color = sect_name, fill = sect_name)) +
  geom_density(alpha = 0.4) +
  coord_flip() +
  theme_void() +
  theme(legend.position = "none")


ggarrange(
  p_x, NULL,
  p_scatter, p_y,
  ncol = 2, nrow = 2,
  widths = c(4, 1),
  heights = c(1, 4),
  common.legend = T,
  legend = "bottom"
)

```

::: notes
Vehicles: Regional GVCs — transport costs, scale economies, regional specialization
Electronics: Global GVCs — highly fragmented production across countries
Agriculture: Less fragmented — stronger local/regional sourcing
Chemicals: Intermediate-input intensive — mainly domestic sourcing, with growing imported contented mostly domestically, but some increasing trend in import content as well
:::

## Four sectors in Japan

```{r}
#| echo: false
#| fig-height: 6
#| fig-width: 12

dt_sec_jpn = fread("../data/03 gvc/gvc_case_jpn_share.csv")

ggbarplot(
  dt_sec_jpn,
  x="sect_name",
  xlab="",
  y="value_sh",
  ylab="percent",
  rotate=T,
  facet.by = "t",
  color="type",
  fill = "type"
) +
  theme_pubclean()
```

## Japan comparative advantage: revisited

```{r}
#| echo: false
dtcaw_jpn = fread("../data/02 ca/rca_jpn.csv")
dtcaw_jpn_s = dtcaw_jpn[t%in%c(1965,2000)]
ggscatter(dtcaw_jpn_s,
          title = "Revealed Comparative Advantage for Japan",
          caption = "Data Source: WITS, WIOD, TIVA",
              color = "sect_group",
          legend="none",
          x="traditional",y="gvc",
          facet.by = "t",
          xlab="Traditional Trade",
          ylab="GVC Trade") +
  geom_abline(intercept = 0, slope = 1, color = "red",
              linetype = "dotdash") +
  geom_hline(yintercept = 1, linetype = "dotdash") +
  geom_vline(xintercept = 1, linetype = "dotdash") +
    geom_text_repel(aes(x = traditional, y = gvc, label = sect_name, 
                      color = sect_group),
                    data = dtcaw_jpn_s, size = 2,
                   show.legend = FALSE)

```
  
::: notes  
- RCA_Ai = share of export of industry i in country A export/ share of global export of industry in global export (Balassa, 1965)
- effective only within same country among products or same product among countries: higher value indicates a stronger export performance, higher degree of export specialization, and a higher degree of revealed comparative advantage
- CA emphasize relative differences in productivity between countries as the reason for international trade
:::  

---
```{r}
#| echo: false
#| fig-height: 9
#| fig-width: 12
dtcaw_reg = fread("../data/02 ca/rca_reg.csv")
regss = unique(dtcaw_reg[,region])
dtcaw_reg[,region:=factor(region,levels = regss)]
dtcaw_reg_s = dtcaw_reg[grepl("Electrical",sect_name)&t%in%c(1970,2000)]
ggscatter(dtcaw_reg_s,
    title = "Revealed Comparative Advantage for Electrical Equipment",
          caption = "Data Source: WITS, TIVA",
          x="traditional",y="gvc",
          #color = "sect_group",
    xlab="Traditional Trade",ylab =" GVC Trade",
          legend="none",
facet.by = "t",nrow=1) +
  geom_abline(intercept = 0, slope = 1, 
              linetype = "dotdash") +
  geom_hline(yintercept = 1, linetype = "dotdash") +
  geom_vline(xintercept = 1, linetype = "dotdash") +
    geom_text_repel(aes(x = traditional, y = gvc, label = region, 
                      #color = sect_group
                      ),
                    data = dtcaw_reg_s, size = 3,max.overlaps = 20,
                   show.legend = FALSE)
```

::: notes  
- the highest degree of RCA variation: primary and unskilled labour-intensive product
- less variation in technology and human capital-intensive product categories
:::  


## Key takeaways

- GVCs fragment production across countries
- GVC participation can take different forms: pure backward, pure forward, and two-sided (mixed) participation.
- Input–output tables make GVC measurement possible by tracing value added through industries and countries across multiple production stages.

---
## Suggested Readings
<div style="font-size:50%">
- Baldwin, R., & Ito, T. (2021). The smile curve: Evolving sources of value added in manufacturing. Canadian Journal of Economics/Revue Canadienne d’économique, 54(4), 1842–1880. 
- Baldwin, R., & Venables, A. J. (2013). Spiders and snakes: Offshoring and agglomeration in the global economy. Journal of International Economics, 90(2), 245–254. https://doi.org/10.1016/j.jinteco.2013.02.005
- Elms, D. K., & Low, P. (2013). Global value chains in a changing world. World trade organization.
- Grossman, G. M., & Rossi-Hansberg, E. (2008). Trading Tasks: A Simple Theory of Offshoring. American Economic Review, 98(5), 1978–1997. https://doi.org/10.1257/aer.98.5.1978
- Koopman, R., Wang, Z., & Wei, S.-J. (2014). Tracing Value-Added and Double Counting in Gross Exports. American Economic Review, 104(2), 459–494. https://doi.org/10.1257/aer.104.2.459
- Kinoshita, T., Kishida, H., & Amemiya, A. (2004). Strategy of Japanese manufacturers in ASEAN and China in light of emerging FTA initiatives.
- OECD. (2011). Globalisation, comparative advantage and the changing dynamics of trade (Online-Ausg). OECD.
- Suvannaphakdy, S., & Pham, T. P. T. (2023). GVC Reconfiguration. ISEAS–Yusof Ishak Institute Singapore.
- Todo, Y., & Inoue, H. (2021). Geographic Diversification of the Supply Chains of Japanese Firms. Asian Economic Policy Review, 16(2), 304–322
- WTO and JETRO-IDE (2011), Trade patterns and global value chains in East Asia: from trade in goods to trade in tasks. 
- Xing, Y., and N. Detert (2010). How the iPhone Widens the United States Trade Deficit with the People’s Republic of China. ADBI Working Paper 257. Tokyo: Asian Development Bank.

## Data
Mancini, M., & Taglioni, D. (2025). Economic consequences of trade and global value chain integration: A measurement perspective. The World Bank Economic Review, https://doi.org/10.1093/wber/lhaf017.

</div>







 
