---
title: "Cleaning Raw Data: A Worked Example"
subtitle: "Maddison Project Database (GDP per capita, 1-2022)"
format: html
---

## The goal of this lesson

This is **not** an R syntax tutorial — it's a walkthrough of a real cleaning
job: turning a messy, human-formatted Excel file into a tidy dataset ready
for analysis and plotting.

The general workflow is always the same, whatever the raw data looks like:

1. **Load** the raw file, one sheet/section at a time
2. **Inspect** it — look at the actual structure before assuming anything
3. **Clean** — rename, reshape, and fix inconsistencies
4. **Handle exceptions** — real data always has special cases
5. **Combine** the cleaned pieces
6. **Sanity-check** with a quick plot before trusting the result
7. **Save** the cleaned data so the raw file is never touched again

We'll follow that exact sequence using the Maddison Project Database (a
long-run historical GDP dataset), which has two of the messiest but most
common problems in real-world data: **data spread across a wide,
label-heavy format**, and **inconsistent structure between sheets**.

## 0. Download the raw data

Before any cleaning can happen, the raw file has to exist locally. The
Maddison Project Database is published on **DataverseNL** with a DOI:
`10.34894/INZBF2`. Downloading it through R (rather than clicking a
button in a browser) means the script documents exactly where the data
came from, and can be re-run from scratch by anyone.

```{r}
#| label: download-raw
#| eval: false
# The dataverse package defaults to Harvard's Dataverse, so it has to be
# pointed at the right server first.
Sys.setenv("DATAVERSE_SERVER" = "dataverse.nl")

# Look up the dataset by its DOI
library(dataverse)
ds <- get_dataset("10.34894/INZBF2")
ds$files[, c("filename", "contentType")]

# Download the specific file by name (confirmed from the listing above)
raw <- get_file("mpd2023_web.xlsx", dataset = "10.34894/INZBF2")

dir.create("../data/raw", showWarnings = FALSE)
dir.create("../data/temp", showWarnings = FALSE)
writeBin(raw, "../data/raw/mpd2023_web.xlsx")
```

::: {.callout-tip}
`ds$files` is worth printing and looking at before downloading anything —
never guess a filename. This is the same "inspect before you act"
habit that shows up again in the cleaning steps below.
:::

This chunk is set to `eval: false` since the file only needs to be
downloaded once. Run it manually the first time, then leave it off for
every subsequent render.

```{r}
#| label: setup
library(readxl)
library(data.table)

mdp <- "../data/raw/mpd2023_web.xlsx"
```

::: {.callout-note}
Keep raw files in a `data/raw/` folder and cleaned output in `data/temp`.
Never overwrite the raw file — if a cleaning step turns out to be wrong,
you want to be able to start over from the original.
:::

---

## 1. Load the raw data

The Maddison file has multiple sheets. We start with the "Regional data"
sheet, which reports GDP per capita and population by world region.

```{r}
#| label: load-region
mdp_region <- read_xlsx(mdp,
                         sheet = "Regional data",
                         skip = 1)
setDT(mdp_region)
```

Note `skip = 1` — the first row of the sheet is a title/header banner, not
column names. **You only know this by opening the file and looking at
it first.** This is a general rule: never write cleaning code before
you've visually inspected the raw file yourself.

---

## 2. Inspect before cleaning

```{r}
#| label: inspect-region
str(mdp_region)
```

This one `str()` call tells us the two problems we need to fix:

- Column names are junk like `...1`, `...10`, `...11` — these come from
  Excel columns that had no header text, or a header that got
  interpreted as blank
- There are **duplicate column names** ("Sub Saharan Africa" appears
  twice — once for GDP per capita, once for population), which R
  silently renamed by appending numbers

Neither of these is visible from the code alone — they only show up once
you actually look at the structure of what got loaded.

---

## 3. Clean: rename ambiguous columns

```{r}
#| label: rename-region
setnames(mdp_region, "...1", "date")
setnames(mdp_region, "...10", "World GDPpc")
setnames(mdp_region, "Sub Saharan Africa", "Sub Saharan Africa___1")
setnames(mdp_region, "Sub Saharan SSA", "Sub Saharan Africa___20")
mdp_region[, "...11" := NULL]
```

Each rename here is fixing a specific, identified problem from step 2 —
not a guess. The two "Sub Saharan Africa" columns get a `___1` /
`___20` suffix as a deliberate marker: it lets us split them back apart
programmatically later, rather than fixing each one by hand.

`mdp_region[, "...11" := NULL]` drops a column that inspection showed was
empty/unused — always confirm a column is genuinely unneeded before
deleting it, rather than dropping anything that looks odd.

---

## 4. Clean: reshape from wide to long

The raw sheet has one column per region — wide format, convenient for a
human reading a spreadsheet, but awkward for analysis. `melt()` converts
it to long format: one row per date/region/value combination.

```{r}
#| label: melt-region
dt <- melt(mdp_region, id.vars = "date")
```

---

## 5. Clean: recover structure from column names

This is the key trick in this dataset. The suffix we added in step 3
(`___1`, `___20`) is now sitting inside the `variable` column, and we can
split it back out into real, usable columns:

```{r}
#| label: split-region
dt[, variable := gsub("\\.", "_", variable)]
dt[, c("region", "id") := tstrsplit(variable, "___")]
dt[, .N, by = id]
```

`dt[, .N, by = id]` here isn't part of the cleaning — it's a **checkpoint**.
Printing a count by group after a split is a cheap way to confirm the
split worked as expected before moving on.

```{r}
#| label: derive-variable-type
dt[, id := as.numeric(id)]
dt[, var := ifelse(id < 10, "GDPpc", "Population")]
dt[, .N, by = .(id, var)]
```

The `id` numbers turned out to encode which variable a region's column
was (GDP per capita vs. population) — this is domain-specific knowledge
you get from reading the file's documentation or reverse-engineering the
pattern, not something obvious from the code.

---

## 6. Handle the exception case

Real datasets almost always have one row/column that doesn't follow the
general pattern. Here, the "World" totals don't have a numeric ID suffix
at all — they were named directly ("World GDPpc", "World Population"),
so the general split rule from step 5 doesn't apply to them:

```{r}
#| label: handle-world
dt[variable %in% c("World GDPpc", "World Population"),
   c("region", "var") := tstrsplit(variable, " ")]
dt[, c("variable", "id") := NULL]
setnames(dt, "var", "variable")
dt[, .N, by = .(variable, region)]
```

Notice the pattern: **apply the general rule, then explicitly patch the
known exception** — rather than writing one rule complex enough to cover
every case at once. This keeps each step readable and debuggable.

---

## 7. Load and clean a second sheet with different structure

Japan isn't broken out separately in the regional sheet, so we pull it
from the "Full data" sheet (which has one row per country-year) instead:

```{r}
#| label: load-japan
mdp_jpn <- read_xlsx(mdp, sheet = "Full data")
setDT(mdp_jpn)
str(mdp_jpn)
```

```{r}
#| label: rename-japan
setnames(mdp_jpn,
         c("year", "gdppc", "pop", "country"),
         c("date", "GDPpc", "Population", "region"),
         skip_absent = TRUE)
```

Renaming to match the column names already used in `dt` (`date`,
`region`) is deliberate — it's what makes the two cleaned pieces
combinable in step 8. `skip_absent = TRUE` means the rename won't error
if a name doesn't exist, which is a safe default when you're not 100%
sure of every column name in a sheet you didn't design.

```{r}
#| label: melt-japan
dt2 <- melt(mdp_jpn[countrycode == "JPN"],
            id.vars = c("date", "region"),
            measure.vars = c("GDPpc", "Population"),
            drop = TRUE)
```

Same wide-to-long reshape as step 4, applied to this differently-shaped
source, filtered down to just the one country we need.

---

## 8. Combine the cleaned pieces

```{r}
#| label: combine
# Real GDP per capita in 2011$
dt3 <- rbindlist(list(dt, dt2), use.names = TRUE, fill = TRUE)
```

`use.names = TRUE` matches columns by name rather than position (safer
whenever two tables were cleaned separately). `fill = TRUE` allows
columns that exist in one piece but not the other — necessary here since
`dt` and `dt2` don't have identical column sets.

The comment above this line matters as much as the code: **always leave
a note about units/definitions right next to the data**, since it won't
be obvious from the column names alone months later.

---

## 9. Sanity-check with a plot before saving

```{r}
#| label: sanity-check-plot
ggpubr::ggline(dt3[variable == "GDPpc" & date > 1900],
               x = "date", y = "value", color = "region",
               numeric.x.axis = TRUE)
```

This plot isn't the final deliverable — it's a **cleaning validation
step**. If a region's line looks discontinuous, jumps to an implausible
value, or is missing entirely, that's a sign of a cleaning bug, and it's
far easier to catch here than after the data has already been saved and
used elsewhere.

::: {.callout-tip}
Keep a commented-out line like this for a second variable you're not
currently checking — it costs nothing and makes it trivial to re-check
later:

```r
# ggpubr::ggline(dt3[variable=="Population"&date>1900],
#                x="date",y="value",color="region",
#                numeric.x.axis = T)
```
:::

---

## 10. Save the cleaned data

```{r}
#| label: save-main
fwrite(dt3[!is.na(value)], "../data/temp/madison.csv")
```

Two decisions worth calling out:

- `dt3[!is.na(value)]` drops rows with no value **at the save step**,
  not earlier — keeping `NA` rows around during cleaning makes it easier
  to spot where gaps came from, but there's no reason to ship them in
  the final file
- Saving to `.csv` in a `data/` (not `data.raw/`) folder means anything
  downstream (a graphing script, another student, future-you) never
  needs to re-open the original Excel file or re-run this cleaning again

---

## 11. Build a companion lookup table

Not every useful output has to be the main dataset. Here we also save a
small reference table mapping country codes to regions, which other
scripts can join against later:

```{r}
#| label: regions-lookup
mdp <- read_xlsx(mdp, sheet = "Full data")
setDT(mdp)
regions <- unique(mdp[, .(countrycode, country, region)])
regions[, .N, by = region]
regions[grepl("Western Offshoots", region), .N, by = country]
fwrite(regions, "../data/temp/madison_regions.csv")
```

The two `.N` checks here are, again, checkpoints — confirming the region
grouping looks sensible (right number of regions, right countries inside
a given region) before trusting the lookup table enough to save it.

---

## Summary: the general pattern

| Step | What | Why |
|---|---|---|
| Load | `read_xlsx()` one sheet at a time | Different sheets often have different structure |
| Inspect | `str()` | Reveals problems (bad names, duplicates) invisible in code alone |
| Rename | `setnames()` | Fix ambiguous/duplicate names *before* reshaping |
| Reshape | `melt()` | Wide (human-readable) → long (analysis-ready) |
| Recover structure | `tstrsplit()` on encoded names | Turn a naming trick back into real columns |
| Handle exceptions | Targeted `dt[condition, ...]` fix | Apply general rule, then patch known special cases |
| Combine | `rbindlist(..., fill = TRUE)` | Merge cleaned pieces from different sources |
| Validate | Quick plot or `.N` by group | Catch cleaning bugs before they reach saved data |
| Save | `fwrite()` to a separate `data/` folder | Never re-touch the raw file; downstream work reads the clean version |

This same sequence applies to almost any messy raw dataset — the details
of *what's* messy change, but the order of operations doesn't.
