---
title: "Cleaning Raw Data: A Worked Example"
subtitle: "WITS - GVC data"
format: html
---

## Overview

This document documents the data-cleaning pipeline for the WITS GVC trade dataset. It covers:

1.  Loading the raw trade, sector, and country reference tables
2.  Classifying countries into analytical regions
3.  Merging trade data with sector and region metadata
4.  Reshaping the merged data into long format for downstream analysis

Analysis, plotting, and index-construction steps are intentionally left out of this document — the goal here is a clean, well-documented, reusable dataset.

```{r setup}
#| message: false
#| warning: false

library(data.table)
library(countrycode)
```

## Download the raw data

All source files live on the World Bank's WITS GVC data page: <https://wits.worldbank.org/gvc/gvc-data-download.html>. Four files are listed there; this pipeline uses three of them:

| File | Link | Size |
|------------------------|------------------------|------------------------|
| GVC Trade data | <https://wits.worldbank.org/data/public/gvc_trade.zip> | \~2.54 GB (zipped) |
| GVC Sector Group and Sectors | <https://wits.worldbank.org/data/public/gvc-sectors.csv> | 14 KB |
| GVC Countries | <https://wits.worldbank.org/data/public/gvc-countries.csv> | 8 KB |

(There is also a *GVC Output* zip on the same page — it is not needed for this trade-cleaning pipeline.)

### Option A — download by hand

1.  Open the page above in a browser.
2.  Click **Download GVC Trade** and save `gvc_trade.zip` into `data.raw/gvc_trade/`.
3.  Unzip it. WITS distributes the trade data as one CSV per source (e.g. `wiodlr`, `tiva`); after unzipping you should see one or more `.csv` files inside `data.raw/gvc_trade/`. Confirm the file name matches what `fread()` expects below (`gvc_trade_WITS_2023.csv`) — rename it if WITS has updated the filename since this pipeline was written.
4.  Click **Download GVC Sector** and **GVC Countries** and save them as `data.raw/gvc-sectors.csv` and `data.raw/gvc-countries.csv` respectively.

### Option B — download from R

The trade zip is large (\~2.5 GB), so this step can take a while and is best run once, outside of routine document rendering.

```{r download-data}
#| eval: false
#| 
#dir.create("../data/raw", showWarnings = FALSE)
gvc_trade = "../data/raw/gvc_trade_WITS_2023.csv"
gvc_sec = "../data/raw/gvc-sectors.csv"
gvc_cty = "../data/raw/gvc-countries.csv"

# GVC trade data (large — ~2.5 GB zipped)
download.file(
  "https://wits.worldbank.org/data/public/gvc_trade.zip",
  destfile = gvc_trade ,
  mode = "wb"
)

# Sector and country reference tables
download.file(
  "https://wits.worldbank.org/data/public/gvc-sectors.csv",
  destfile = gvc_sec , mode = "wb"
)
download.file(
  "https://wits.worldbank.org/data/public/gvc-countries.csv",
  destfile = gvc_cty, mode = "wb"
)
```

After unzipping, list the contents to confirm the trade CSV's exact name before moving on — WITS may package it differently than expected:

```{r list-unzipped}
#| eval: false

list.files("../data/raw")
```

## Load raw data

Three source files are needed:

- `dt` — the raw bilateral trade / GVC flows (millions of USD)
- `sectors` — sector code-to-group crosswalk
- `cty` — country code reference table

```{r load-data}
dt      = fread("../data/raw/gvc_trade_WITS_2023.csv")
sectors = fread("../data/raw/gvc-sectors.csv")
cty     = fread("../data/raw/gvc-countries.csv")
```

```{r inspect-raw}
str(dt)
head(dt)
```

## Classify countries into regions

We build a `Region2` grouping on top of the country table:

- **Europe** — all countries in the UN "Europe" region
- **ASEAN** — countries in the UN "South-eastern Asia" sub-region
- **N.America** — `USA`, `CAN`, `MEX` (NAFTA)
- Individually tracked countries: `JPN`, `CHN`, `KOR`
- Everything else falls into **ROW** (rest of world)

```{r region-groups}
eurs  = codelist$iso3c[codelist$un.region.name == "Europe" &
                          !is.na(codelist$iso3c)]

asean = codelist$iso3c[codelist$un.regionsub.name == "South-eastern Asia" &
                          !is.na(codelist$iso3c)]

nafta = c("USA", "CAN", "MEX")

idvs  = c("JPN", "CHN", "KOR")

cty[, Region2 := country]
cty[country %in% eurs,  Region2 := "Europe"]
cty[country %in% asean, Region2 := "ASEAN"]
cty[country %in% nafta, Region2 := "N.America"]
cty[!country %in% c(eurs, asean, nafta, idvs), Region2 := "ROW"]
```

### Sanity check: any countries without a region?

```{r region-check}
cty[is.na(Region2)]
```

If this returns rows, those countries were not matched by `countrycode`'s region lookup and should be assigned manually before proceeding.

## Merge trade data with sector groupings

Trade is merged onto the sector crosswalk using `sect` and `source` (the data source, e.g. `wiodlr` or `tiva`), since sector codes are not necessarily consistent across sources.

```{r merge-sectors}
dt2 = merge(dt, sectors, by = c("sect", "source"))
```

### Sanity check: sector coverage by source

```{r sector-check}
dt2[, .N, by = .(source, sect_group)][order(source, sect_group)]

dt2[source == "wiodlr", .N, by = .(sect_group, sect_name)][order(sect_group)]
```

## Merge in exporter / importer regions

The region classification built above is joined twice — once for the exporter (`exp`) and once for the importer (`imp`).

```{r merge-regions}
dt2 = merge(dt2, cty[, .(exp = country, exp_reg = Region2)], by = "exp")
dt2 = merge(dt2, cty[, .(imp = country, imp_reg = Region2)], by = "imp")
```

```{r merged-check}
dim(dt2)
sum(is.na(dt2$exp_reg))
sum(is.na(dt2$imp_reg))
```

Both `NA` counts should be zero after this step — any non-zero count means some exporter/importer codes in `dt` don't have a match in `cty` and were dropped by the merge (or need to be added to the country table).

## Reshape to long format

Two sources are combined:

- `wiodlr` for all years
- `tiva` restricted to years after 2000 (its methodology changes before that point)

The GVC/trade-type columns (`traditional_trade_int`, `traditional_trade_fin`, `gvcbp`, `gvcfp`, `gvcmix`) are melted into a single `variable` / `value` pair, which is the tidy shape most of the downstream analysis expects.

```{r reshape-long}
dtl = melt(
  dt2[source == "wiodlr" | (source == "tiva" & t > 2000),
      .(source,imp, exp, imp_reg, exp_reg,
        sect_name, sect_group,
        t, traditional_trade_int, traditional_trade_fin,
        gvcbp, gvcfp, gvcmix)],
  id.vars = c("imp", "exp",
              "imp_reg", "exp_reg",
              "sect_name", "sect_group", "t")
)
```

```{r reshape-check}
dtl[, .N, by = variable]
range(dtl$t)
dt[,.N,by=source]
```

## Save the cleaned dataset

Persist the cleaned, merged, long-format table so downstream scripts don't need to repeat the loading/merging steps above. Use parquet file to reduce on file size.

```{r save-clean}
#| eval: false

arrow::write_parquet(dtl, "../data/temp/gvc_trade.parquet")
```

## Summary

| Step | Object | Description |
|------------------------|------------------------|------------------------|
| Load | `dt`, `sectors`, `cty` | Raw trade, sector, and country tables |
| Classify | `cty$Region2` | Country → region mapping (Europe, ASEAN, N.America, JPN/CHN/KOR, ROW) |
| Merge sectors | `dt2` | Trade + sector group, joined on `sect` + `source` |
| Merge regions | `dt2` | Adds `exp_reg` and `imp_reg` |
| Reshape | `dtl` | Long-format trade table (`wiodlr` all years + `tiva` post-2000) |

`dtl` is the base table that all subsequent analysis (regional shares, GVC participation indices, revealed comparative advantage, network graphs, etc.) should build on.