---
title:  "Cleaning Raw Data: A Worked Example"
subtitle: "Penn World table: Growth Accounting"
format: html
---

## Download the data

The `PWT_GA.csv` file used below is the **Growth Accounting (GA)** query from
Penn World Table 11.0. GGDC does not publish a single static CSV link for each
query, so the file has to be exported once from their online tool and then
read locally (or you can download the full dataset programmatically instead).

**Option A — GA query via the online tool (recommended for this script)**

1. Go to the [PWT online data access tool](https://pwt-data-tool.streamlit.app/?page=GA).
2. Select the countries/years you want (or leave the defaults for all countries).
3. Export/download the result as CSV.
4. Save it as `data.raw/PWT_GA.csv` (the path the script below expects).

**Option B — download the full PWT release programmatically**

If you'd rather work from the full dataset and derive the GA variables
yourself, you can pull the Excel file directly from Dataverse:

```{r}
#| eval: false
dir.create("../data/raw", showWarnings = FALSE)

download.file(
  url = "https://dataverse.nl/api/access/datafile/554105",
  destfile = "data/raw/pwt110.xlsx",
  mode = "wb"
)
```

Check the [PWT 11.0 page](https://www.rug.nl/ggdc/productivity/pwt/) for the
current release and file IDs, since these links are updated with each new
version.

## Load data

```{r}
library(data.table)

pwt = "../data/raw/PWT_GA.csv"

pwtd = fread(pwt, header = T)
setnames(pwtd, gsub(" ", "", colnames(pwtd)))
str(pwtd)
pwtd[, .N, by = .(Variablecode, Variablename)]
```

## Reshape to long format

```{r}
pwtdl = melt(pwtd,
             id.vars = c("Variablecode", "Variablename",
                         "ISOcode", "Country"),
             variable.name = "date",
             variable.factor = FALSE)

setorder(pwtdl, ISOcode, Variablename, date)
```

## Compute year-over-year growth

```{r}
pwtdl[, Lvalue := shift(value, type = "lag"),
      by = .(ISOcode, Variablename)]
pwtdl[, gyoy := value / Lvalue - 1]
pwtdl[, date := as.numeric(date)]

pwtdl[, Variablename2 := trimws(gsub("(\\(|\\,|by\\b|at\\b).*$", "",
                                      Variablename))]
```

## Plot growth rates (Japan vs US)

```{r}
ggpubr::ggline(pwtdl[ISOcode %in% c("JPN", "USA")
                      & !grepl("Share", Variablename)
                     &!is.na(gyoy)],
               x = "date", y = "gyoy", color = "ISOcode",
               facet.by = "Variablename2",
               numeric.x.axis = T)
```

## Plot index variables

```{r}
ggpubr::ggline(pwtdl[ISOcode %in% c("JPN", "USA")
                      & grepl("index", Variablename)],
               x = "date", y = "value", color = "ISOcode",
               facet.by = "Variablename2",
               numeric.x.axis = T)
```

## Plot share variables

```{r}
ggpubr::ggline(pwtdl[ISOcode %in% c("JPN", "USA")
                      & grepl("Share", Variablename)],
               x = "date", y = "value", color = "ISOcode",
               facet.by = "Variablename2",
               numeric.x.axis = T)
```

## Save processed data

```{r}
fwrite(pwtdl[, .(Variablecode,
                  Variablename2,
                  ISOcode, date,
                  value, gyoy)],
       "../data/temp/pwt.csv")
```
